Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-20 paper/numresults.tex
polish results and add conclusion

In this last section, we test our algorithm in multiple scenarios. First, we numerically confirm our theoretical convergence bounds by approximating the spectral density of a Hamiltonian arising from electronic structure interaction. Subsequently, we compare our method to the Krylov-aware stochastic trace estimator \cite{chen-2023-krylovaware-stochastic} once on the electronic structure example and once for estimating the trace of a more general parameter-dependent matrix appearing in statistical thermodynamics. In the end, we demonstrate that our algorithm can also be used to monitor the optimization process of a neural network.

@@

Our first numerical example comes from electronic structure interaction \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
\begin{equation}
    v(r) = v_0 e^{-\lambda r^2}
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf. \reffig{fig:gaussian-well}). The discretization step is fixed to be $h=0.6$, such that the size of the matrix grows cubically with $n_c$. This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian -- its spectral density -- allows us to interpret the system's energy levels.

@@

While theoretically an important tool, we waive using the non-negative Chebyshev expansion, since the slight indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-expansion} does not make a difference in finite precision arithmetic. We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using a midpoint quadrature with $n_t \in \mathbb{N}$ nodes. For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}).

\begin{figure}[ht]
    \centering
    \input{plots/convergence.pgf}
    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$ we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$. The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes.}
    \label{fig:convergence}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \input{plots/distribution.pgf}
    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter. We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
    \label{fig:distribution}
\end{figure}

@@

To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters, which we train on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.}. We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss. It can well be observed that the eigenvalues creep towards zero as the training proceeds. Furthermore, by the absence of large eigenvalues in the spectrum, we can tell that in the final epoch, the neural network attains a more robust state than in previous epochs.

>>>>

\color{black}

In this last section, we test our algorithm in multiple scenarios. First, we numerically confirm our theoretical convergence bounds by approximating the spectral density of a Hamiltonian arising from electronic structure interaction. Subsequently, we compare our method to the Krylov-aware stochastic trace estimator, once on the electronic structure example, and once for estimating the trace of a more general parameter-dependent matrix appearing in statistical thermodynamics. In the end, we demonstrate that our algorithm can also be used to monitor the optimization process of a neural network.

All our implementations are written from scratch in Python 3.12.3 using the packages NumPy 2.0.0 and SciPy 1.14.1. They are executed on a single thread of a GitHub-hosted Ubuntu runner with a 64-bit processor and 16 GB of RAM. \input{re-pro-badge.tex}

@@

Our first numerical example comes from electronic structure interaction \cite{lin-2017-randomized-estimation}, more precisely from the second order finite difference discretization of the Hamiltonian
\begin{equation}
    \mathcal{H} = - \Delta + V
    \label{equ:5-experiments-electronic-hamiltonian}
\end{equation}
in three dimensions. The potential $V$ interacting with the electrons is generated by Gaussian wells
\begin{equation}
    v(r) = v_0 e^{-\lambda r^2}
    \label{equ:5-experiments-gaussian-cell}
\end{equation}
with $v_0 = -4$ and $\lambda = 8$, centered in cells of side length $L=6$ which are stacked $n_c \in \mathbb{N}$ times in each spatial dimension (cf. \reffig{fig:gaussian-well}). The discretization step is fixed to be $h=0.6$, such that the size of the matrix grows cubically with $n_c$. This is an idealized model for the interaction of nuclei on a regular grid with electrons for a $k$-vector in the center of the first Brillouin zone. The distribution of the eigenvalues of the Hamiltonian --- its spectral density --- allows us to interpret the system's energy levels.

@@

While theoretically an important tool, for the remainder of this section, we waive using the non-negative Chebyshev approximation, since the (slight) indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not seem to matter when computing the Nyström++ estimator in finite precision arithmetic. We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using a midpoint quadrature with $n_t \in \mathbb{N}$ nodes. For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}).

\begin{figure}[ht]
    \centering
    \input{plots/convergence.pgf}
    \caption{For increasing values of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$ but fixed $m$ in the Chebyshev-Nyström++ method, we plot the $L^1$-approximation error for $\sigma=0.005$ on the model problem with $n_c = 1$. The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes in $[-1, 1]$.}
    \label{fig:convergence}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \input{plots/distribution.pgf}
    \caption{The Chebyshev-Nyström++ method is applied to the model problem with $n_c = 1$ for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter. We make the approximation error made in the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}). The $L^1$-error is approximated by a quadrature with $n_t = 100$ evenly spaced nodes in $[-1, 1]$.}
    \label{fig:distribution}
\end{figure}

@@

To demonstrate that our algorithm can effectively be applied in the setting of neural network optimization, we approximate the spectral density of a small fully connected convolutional neural network with 6782 parameters. We train this network on the handwritten digit classification task given by the MNIST dataset\footnote{Handwritten digit classification; taken from \url{http://yann.lecun.com/exdb/mnist}.} in PyTorch 2.4.1 in a standard way. We plot the approximation of the spectral density of the Hessian matrix of the untrained neural network, and at different stages of training in \reffig{fig:hessian-density}, as well as the corresponding mean squared error loss. It can well be observed that the eigenvalues creep towards zero as the training proceeds. Furthermore, despite the loss steadily decreasing, the presence of large eigenvalues in some epochs indicates a sharp minimum of the network, hence, unfavorable generalization properties.

### 2025-02-20 paper/spectraldensity.tex
polish results and add conclusion

The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care. It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantum}. Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver which likewise uses a truncation of the smallest eigenvalues to enforce better conditioning.

What we have found to be crucial for a successful approximation of the full spectral density was the detection of regions wherein lie no eigenvalues, i.e. regions where the spectral density nearly vanishes. For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}_1(t) \approx \mtx{\Omega}^{\top} g_{\sigma}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$ to be close to the zero matrix for parameter values $t$ located in such regions. Hence, on \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable (see \reffig{fig:zerocheck}). Notice that with $\mtx{K}_1(t)$ -- a matrix we need to assemble anyway -- we can cheaply compute $\Trace(\mtx{K}_1(t)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator. We may use it as a rough indicator whether $t$ lies within a region where the spectral density nearly vanishes, and subsequently set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.

\begin{figure}[ht]
    \centering
    \input{plots/zerocheck.pgf}
    \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}. Here, we used a matrix described in \refsec{sec:results} and ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse. We let $m=2000$, $n_{\mtx{\Omega}}=80$, $n_{\mtx{\Psi}}=0$, and $\sigma = 0.004$.}
    \label{fig:zerocheck}
\end{figure}

>>>>

The pseudoinverses on \reflin{lin:4-nystromchebyshev-nystrom-pp} need to be computed with care. It is possible to reformulate the problem as a generalized eigenvalue problem and apply thresholding to the smallest eigenvalues of a certain matrix to enforce stability \cite{lin-2017-randomized-estimation, epperly-2022-theory-quantum}. Since this renders the algorithm significantly more complicated but only improved the accuracy slightly during our numerical experiments, we will instead evaluate these expressions in \reflin{lin:4-nystromchebyshev-nystrom-pp} using a least-squares solver, which likewise uses a truncates the smallest eigenvalues by treating them as zero if they are smaller than $10^{-5}$ times the largest eigenvalue.

What we have found to be crucial for a successful approximation of the full spectral density was the detection of regions wherein lie no eigenvalues, i.e. regions where the spectral density nearly vanishes. For small smoothing parameters $\sigma$, the rapid decay of the Gaussian $g_{\sigma}$ will cause $\mtx{K}_1(t) \approx \mtx{\Omega}^{\top} g_{\sigma}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$ to be close to the zero matrix for parameter values $t$ located in such regions. Hence, on \reflin{lin:4-nystromchebyshev-nystrom-pp} we would compute the pseudoinverse of an almost zero matrix, which can be extremely unstable (see \reffig{fig:zerocheck}). Notice that with $\mtx{K}_1(t)$ --- a matrix we need to assemble anyway --- we can cheaply compute $\Trace(\mtx{K}_1(t)) / n_{\mtx{\Omega}}$, which is the $n_{\mtx{\Omega}}$-query Girard-Hutchinson estimator \refequ{equ:hutchinson-trace-estimator}. If this estimate is significantly smaller than $1/n$, in our experiments usually $10^{-5}$, we may use it as an indicator that $t$ lies within a region where the spectral density nearly vanishes, and therefore set $\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = 0$ instead of computing the full expression on \reflin{lin:4-nystromchebyshev-nystrom-pp}.

\begin{figure}[ht]
    \centering
    \input{plots/zerocheck.pgf}
    \caption{The difference the non-zero check can make when approximating a spectral density using \refalg{alg:nystrom-chebyshev-pp}. Here, we use a coarse discretization of the Hamiltonian \refequ{equ:5-experiments-electronic-hamiltonian} in three dimensions and ran the Chebyshev-Nyström++ method with and without estimating if the matrix function is zero before taking its pseudo-inverse. We let $m=2000$, $n_{\mtx{\Omega}}=80$, $n_{\mtx{\Psi}}=0$, and $\sigma = 0.004$. We compute the $L^1$ error from the approximation of the spectral density at $n_t = 100$ uniformly spaced values of $t$ in the interval $[-1, 1]$.}
    \label{fig:zerocheck}
\end{figure}
